---
title: Invited Talk at University of San Francisco, February 2020
author: Greg Dubrow
date: '2021-02-26'
slug: invited-talk-at-university-of-san-francisco-february-2020
draft: true
categories:
  - post
tags:
  - data visualization
  - ggplot
  - higher eduction
subtitle: ''
summary: ''
authors: []
lastmod: '2021-02-01T11:26:23-08:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

Back in early 2020 I was invited to give a talk at two budget forums at the [University of San Francisco](https://www.usfca.edu/). The general theme was looking at the landscape of enrollments in higher education, with a specific focus on liberal arts colleges, especially Jesuit colleges. Because I collected data from a variety of sources I thought it was make for a good blog post. Plus it's the first time I'm posting about higher education, which has been my area of professional expertise for a while now.

I structured the talk around these general questions:

### What are the major trends affecting the higher education landscape in the US today, particularly traditional liberal arts colleges?

  * Changing demographics impacting enrollments
  * Affordability  
  * What social/political threats are on the horizon that colleges should start addressing?

The talk was divided into four segments:

  * The high school graduation picture in California (USF gets most of its students from CA)
  * Historical enrollment at USF, and compared to other Jesuit colleges
  * College Costs and Affordability  
  * What social/political threats are on the horizon that colleges should start addressing?

The data, code and resulting presentation are in this [github repo](https://github.com/greg-dubrow/USF-talk-February-2020). What I plan to do in this post is in effect annotate some of the code to explain how I put everything together.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "images/",
  out.width = "100%") 
```

First up is pulling in a few decades of high school graduation and enrollment data and wrangling it all to show historical enrollment and projections through 2029. The full code for that is at the github repo in the file [01_hs enrollment data.R](https://github.com/greg-dubrow/USF-talk-February-2020/blob/master/01_hs%20enrollment%20data.R). So what did I do?

First, loaded some packages:
```{r pkg load, message=FALSE, ECHO = TRUE, collapse= TRUE, eval= FALSE}
# readin in data, create df for plots
library(tidyverse) # to do tidyverse things
library(tidylog) # to get a log of what's happening to the data
library(rCAEDDATA) # aggregated CA HS data
library(readr) # read in flat files
library(janitor) # data munging/cleaning utilities
```

The [rCAEDDATA package](https://github.com/daranzolin/rCAEDDATA) was put together by my former [SFSU IR](https://ir.sfsu.edu/) colleague [David Ranzolin](https://daranzolin.github.io/). Though the last update was October 2017, it still contains a trove of data, including public HS graduation numbers from 1993 to 2016. That said, I didn't actually use the package functions, just downloaded the data included with the package. For later years I manually downloaded files from the CA Department of Education's data pages: [2017 here](https://www.cde.ca.gov/ds/sd/sd/filesgrad.asp), and the [later years here](https://www.cde.ca.gov/ds/sd/sd/filesfycgr.asp). The 2017 file has the same structure as the files in David's package. Later years need a bit of restructuring:
```{r hsgrad1, message=FALSE, ECHO = TRUE, collapse= TRUE, eval= FALSE, tidy=TRUE}
cahsgrad18 <- read.delim("data/cahsgrad18.txt", stringsAsFactors=FALSE) %>%
	clean_names() %>%
	filter(reporting_category == "TA") %>%
	filter(aggregate_level == "S") %>%
	filter(dass == "All") %>%
	filter(charter_school == "All") %>%
	# filter(school_name != "Nonpublic, Nonsectarian Schools") %>%
	# filter(school_name != "District Office") %>%
	mutate(YEAR = "2018") %>%
	mutate(YEAR = factor(YEAR)) %>%
	mutate_at(vars(ends_with("_code")), as.character) %>%
	mutate(county_code = ifelse(nchar(county_code) == 1, str_pad(county_code, 2, "left", "0"), county_code)) %>%
	mutate(CDS_CODE = paste(county_code, district_code, school_code, sep = "")) %>%
	mutate(GRADS = as.integer(ifelse(regular_hs_diploma_graduates_count == "*",
																	 0, regular_hs_diploma_graduates_count))) %>%
	mutate(UC_GRADS = as.integer(ifelse(met_uc_csu_grad_req_s_count == "*",
																		 0, met_uc_csu_grad_req_s_count))) %>%
	select(CDS_CODE, GRADS, UC_GRADS, YEAR) %>%
	group_by(YEAR) %>%
	summarise(total_grads = sum(GRADS),
						uccsu = sum(UC_GRADS),
						notuccsu = total_grads - uccsu)

```
Next, some projected HS graduation data from the [CA Department of Finance](https://www.dof.ca.gov/Forecasting/Demographics/Projections/Public_K-12_Graded_Enrollment/). I did a quick transposing of the "HS Grads Table" tab in excel and used that to read into r. You can see the file I used at the github repo's [data folder](https://github.com/greg-dubrow/USF-talk-February-2020/tree/master/data).
```
grproj_to2028 <- readxl::read_excel("data/capublic_k12_enrollproj_to2028.xlsx",
																						 sheet = "hsgrads-tr") %>%
	filter(year != "2017-18") %>%
	mutate(yearend = str_sub(year, 6, 7)) %>%
	mutate(YEAR = paste("20", yearend, sep = "")) %>%
	#mutate(YEAR = factor(year_ch)) %>%
	mutate(uccsu = as.integer(NA)) %>%
	mutate(notuccsu = as.integer(NA)) %>%
	mutate(notuccsu = as.integer(NA)) %>%
	mutate(type = "Projected") %>%
	select(YEAR, total_grads = total, uccsu, notuccsu, type) %>%
	# amend 2018-19 with actual results from
	# https://dq.cde.ca.gov/dataquest/dqcensus/CohRateLevels.aspx?cds=00&agglevel=state&year=2018-19
	mutate(total_grads = ifelse(YEAR == "2019", 417496, total_grads)) %>%
	mutate(uccsu = ifelse(YEAR == "2019", 210980, uccsu)) %>%
	mutate(notuccsu = ifelse(YEAR == "2019", total_grads - uccsu, notuccsu)) %>%
	mutate(type = ifelse(YEAR == "2019", "Actual", type))
```

After cleaning just rbind the files together and voila...actual & projected HS graduation data from 1993 to 2029!


